A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-17 arxiv/proof_examples.tex
pf

{\cor(start here)} Note that the condition $\ex{\norm{\beta_1 - \beta_2}^2}< \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$ for $\rho_1 > 40$ gives that $\delta_{\vari}>\delta_{\bias} $ by Proposition \ref{prop_dist_transition}. Hence if $\sigma_1^2\le \sigma_2^2$, then $\delta_{\vari}^{(2)}<0$ and we always have $\delta_{\vari} > \delta_{\bias}+ \delta_{\vari}^{(2)}$, which gives $\te(\hat{\beta}_2^{\MTL})<\te(\hat{\beta}_2^{\STL})$.  It remains to consider the case $\sigma_1^2 \ge \sigma_2^2$.

\noindent{\bf Positive transfer.} By \eqref{Deltavar}, \eqref{Deltabeta} and \eqref{Deltavar2}, if the following inequality holds,
\be\label{cond sigma1}
\begin{split} 
&\sigma_2^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)} \cdot (1-\oo(1)) \\
&>pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\al_+^2(\rho_1)  + (\sigma_1^2 -\sigma_2^2)\cdot \frac{\rho_1 (\rho_1+\rho_2)}{(\rho_1+\rho_2-1)^3} \al_+(\rho_1)  , 
\end{split}
\ee
then we have $\delta_{\vari} > \delta_{\bias} + \delta_{\vari}^{(2)}$ w.h.p., which gives $\te(\hat{\beta}_t^{\MTL})<\te(\hat{\beta}_t^{\STL})$. We can solve \eqref{cond sigma1} to get
\begin{align*}
\sigma_1^2 < - pd^2\cdot \rho_1 \al_+(\rho_1) +\sigma_2^2 \left[ 1+ \rho_1\Phi(\rho_1, \rho_2) \al_+^{-1}(\rho_1)\right]\cdot (1-\oo(1)).
\end{align*}
This proves the first claim of Proposition \ref{prop_var_transition} using $\nu\ge \al_+^2(\rho_1) +  \oo(1) $.

@@

Next we state Proposition \ref{prop_data_efficiency}, which gives precise upper and lower bounds on the data efficiency ratio for Taskonomy. %In the statement, we shall denote the data efficiency ratio as $\al^\star$.

\begin{proposition}%[Labeled data efficiency]
\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1,\rho_2 \ge 9$ and $\Psi(\beta_1, \beta_2) < (5(\rho_1-1))^{-1} + (5(\rho_2-1))^{-1}$.
	Then the data efficiency ratio $x^\star$ satisfies 
	\be\label{eq_uplowx} x_l  \le x^\star\le \frac{1}{\rho_1 + \rho_2} \bigbrace{  \frac{2}{(\rho_1-1)^{-1} + (\rho_2-1)^{-1} - 5\Psi(\beta_1, \beta_2)}+1}, \ee
	where we denoted
	$$x_l:= \frac1{\rho_1+\rho_2}\left(\frac{2}{(\rho_1-1)^{-1}+(\rho_2 -1)^{-1}}+1\right).$$
\end{proposition}

\begin{proof}%[Proof of Proposition \ref{prop_data_efficiency}]
Suppose we have reduced number of datapoints---$x n_1$ for task 1 and $x n_2$ for task 2 with $n_1=\rho_1 p$ and $n_2=\rho_2 p$. Then all the results in the proof of Proposition \ref{prop_dist_transition} still hold, except that we need to replace $(\rho_1,\rho_2)$ with $(x\rho_1,x\rho_2)$. More precisely, we have
	\begin{align*}
		 a_1 = \frac{\rho_1(x\rho_1 + x\rho_2 - 1)}{x(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(x\rho_1 + x\rho_2 - 1)}{x(\rho_1 + \rho_2)^2} ,  \\
		 a_3 = \frac{\rho_2}{(\rho_1 + \rho_2)(x\rho_1 + x\rho_2 - 1)}, \quad
		& a_4 = \frac{\rho_1}{(\rho_1 + \rho_2)(x\rho_1 + x\rho_2 - 1)}. 
	\end{align*}
Moreover, with high probability,
	\be\label{reduceproof1}
\begin{split}
\te_i(\hat \beta_i^{\MTL}(x)) =  \frac{\sigma^2}{x (\rho_1+\rho_2) - 1}\left(1+ \oo(1)\right)+ \delta^{(i)}_{\bias}, \quad i=1,2.
\end{split}
\ee
 Here the model shift biases $\delta^{(i)}_{\bias}$ satisfy that  
\be\nonumber %\label{Deltabeta} 
\al_-^2(\al\rho_i) - \oo(1)  \le {\delta_{\bias}^{(i)}}/{ \Delta_{\bias}^{(i)}} \le \al_+^2(\al\rho_i) +  \oo(1) , \quad i=1,2,\ee	
where $ \Delta^{(i)}_{\bias}$ are defined as 
\be \nonumber
\Delta^{(i)}_{\bias} := pd^2 \frac{(x\rho_i)^2\cdot x (\rho_1+\rho_2)}{[x ( \rho_1+\rho_2) - 1]^3} ,\quad i=1,2,. 
\ee
On the other hand, using Lemma \ref{lem_minv} we have w.h.p.,
\be\label{reduceproof2} 
\te_i(\hat{\beta}_i^{\STL}) = \frac{\sigma^2}{\rho_i -1} \left( 1+ \oo(1)\right),\quad i=1,2.
\ee
Comparing \eqref{reduceproof1} and \eqref{reduceproof2}, we immediately obtain the lower bound $x^\star\ge x_l $. 
%where
%$$\al_l:=\frac1{\rho_1+\rho_2}\left(\frac{2}{\frac1{\rho_1-1}+\frac1{\rho_2 -1}}+1\right) \ge \frac{\min(\rho_1,\rho_2)}{\rho_1+\rho_2}.$$
In fact, one can see that if $x< x_l$, then we have 
$$ \frac{2\sigma^2}{x (\rho_1+\rho_2) - 1} > \frac{\sigma^2}{\rho_1-1}+\frac{\sigma^2}{\rho_2-1},$$
that is, $\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))$ is larger than $\te_1(\hat{\beta}_t^{\STL}) + \te_2(\hat{\beta}_t^{\STL})$ even if we do not take into account the model shift bias terms $ \delta^{(i)}_{\bias}$.

>>>>

 Note that the condition $pd^2< \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$ for $\rho_1 > 40$ gives that $\delta_{\vari}>\delta_{\bias} $ by Proposition \ref{prop_dist_transition}. Hence if $\sigma_1^2\le \sigma_2^2$, then $\delta_{\vari}^{(2)}<0$ and we always have $\delta_{\vari} > \delta_{\bias}+ \delta_{\vari}^{(2)}$, which gives $\te(\hat{\beta}_2^{\MTL})<\te(\hat{\beta}_2^{\STL})$.  It remains to consider the case $\sigma_1^2 \ge \sigma_2^2$.

\noindent{\bf Positive transfer.} By \eqref{Deltavar}, \eqref{Deltabeta} and \eqref{Deltavar2}, if the following inequality holds,
\be\label{cond sigma1}
\begin{split} 
&\sigma_2^2  \cdot \frac{\rho_1}{(\rho_2-1)(\rho_1 + \rho_2 -1)} \cdot (1-\oo(1)) \\
&>pd^2 \cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\al_+^2(\rho_1)  + (\sigma_1^2 -\sigma_2^2)\cdot \frac{\rho_1 (\rho_1+\rho_2)}{(\rho_1+\rho_2-1)^3} \al_+(\rho_1)  , 
\end{split}
\ee
then we have $\delta_{\vari} > \delta_{\bias} + \delta_{\vari}^{(2)}$ w.h.p., which gives $\te(\hat{\beta}_2^{\MTL})<\te(\hat{\beta}_2^{\STL})$. We can solve \eqref{cond sigma1} to get
\begin{align*}
\sigma_1^2 < - pd^2\cdot \rho_1 \al_+(\rho_1) +\sigma_2^2 \left[ 1+ \rho_1\Phi(\rho_1, \rho_2) \al_+^{-1}(\rho_1)\right]\cdot (1-\oo(1)).
\end{align*}
This proves the first claim of Proposition \ref{prop_var_transition} using $\nu\ge \al_+^2(\rho_1) +  \oo(1) $.

@@

Next we prove Proposition \ref{prop_data_efficiency}, which gives upper and lower bounds on the data efficiency ratio for Taskonomy.

\begin{proof}[Proof of Proposition \ref{prop_data_efficiency}]
Suppose we have reduced number of datapoints---$x n_1$ for task 1 and $x n_2$ for task 2 with $n_1=\rho_1 p$ and $n_2=\rho_2 p$. Then all the results in the proof of Proposition \ref{prop_dist_transition} still hold, except that we need to replace $(\rho_1,\rho_2)$ with $(x\rho_1,x\rho_2)$. More precisely, we have
	\begin{align*}
		 a_1 = \frac{\rho_1(x\rho_1 + x\rho_2 - 1)}{x(\rho_1 + \rho_2)^2} ,\quad
		& a_2 = \frac{\rho_2(x\rho_1 + x\rho_2 - 1)}{x(\rho_1 + \rho_2)^2} ,  \\
		 a_3 = \frac{\rho_2}{(\rho_1 + \rho_2)(x\rho_1 + x\rho_2 - 1)}, \quad
		& a_4 = \frac{\rho_1}{(\rho_1 + \rho_2)(x\rho_1 + x\rho_2 - 1)}. 
	\end{align*}
Moreover, with high probability,
	\be\label{reduceproof1}
\begin{split}
\te_i(\hat \beta_i^{\MTL}(x)) =  \frac{\sigma^2}{x (\rho_1+\rho_2) - 1}\left(1+ \oo(1)\right)+ \delta^{(i)}_{\bias}, \quad i=1,2.
\end{split}
\ee
 Here the model shift biases $\delta^{(i)}_{\bias}$ satisfy that  
\be\nonumber %\label{Deltabeta} 
\al_-^2(\al\rho_i) - \oo(1)  \le {\delta_{\bias}^{(i)}}/{ \Delta_{\bias}^{(i)}} \le \al_+^2(\al\rho_i) +  \oo(1) , \quad i=1,2,\ee	
where $ \Delta^{(i)}_{\bias}$ are defined as 
\be \nonumber
\Delta^{(i)}_{\bias} := pd^2 \frac{(x\rho_i)^2\cdot x (\rho_1+\rho_2)}{[x ( \rho_1+\rho_2) - 1]^3} ,\quad i=1,2,. 
\ee
On the other hand, using Lemma \ref{lem_minv} we have w.h.p.,
\be\label{reduceproof2} 
\te_i(\hat{\beta}_i^{\STL}) = \frac{\sigma^2}{\rho_i -1} \left( 1+ \oo(1)\right),\quad i=1,2.
\ee
Comparing \eqref{reduceproof1} and \eqref{reduceproof2}, we immediately obtain the lower bound $x^\star\ge x_l $. 
%where
%$$\al_l:=\frac1{\rho_1+\rho_2}\left(\frac{2}{\frac1{\rho_1-1}+\frac1{\rho_2 -1}}+1\right) \ge \frac{\min(\rho_1,\rho_2)}{\rho_1+\rho_2}.$$
In fact, one can see that if $x< x_l$, then we have 
$$ \frac{2\sigma^2}{x (\rho_1+\rho_2) - 1} > \frac{\sigma^2}{\rho_1-1}+\frac{\sigma^2}{\rho_2-1},$$
that is, $\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))$ is larger than $\te_1(\hat{\beta}_t^{\STL}) + \te_2(\hat{\beta}_t^{\STL})$ even if we do not take into account the model shift bias terms $ \delta^{(i)}_{\bias}$.

### 2020-08-17 arxiv/transfer.tex
sec 2 wip
pf

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}%
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.
The result can be found in Proposition \ref{prop_data_efficiency} in Appendix \ref{app_proof_32}.

>>>>

\textbf{Connection to Taskonomy.} We use our tools to explain a key result of Taskonomy by Zamir et al. \cite{ZSSGM18}, which shows that MTL can reduce the amount of labeled data needed to achieve comparable performance to STL.
%More precisely, suppose we have $n_i$ datapoints for each task, for $i= 1, 2$.
For $i = 1, 2$, let $\hat{\beta}_i^{\MTL}(x)$ denote the estimator trained using $x \cdot n_i$ datapoints from every task. The data efficiency ratio is defined as
{\small\[ \argmin_{x\in(0, 1)} ~~
		\te_1(\hat{\beta}_1^{\MTL}(x)) + \te_2(\hat{\beta}_2^{\MTL}(x))
		\le \te_1(\hat{\beta}_1^{\STL}) + \te_2(\hat{\beta}_2^{\STL}). \]}
For example, the data efficiency ratio is $1$ if there is negative transfer.
Using our tools, we show that in the isotropic model, the data efficiency ratio is
roughly
{\small\[ \frac{1}{\rho_1 + \rho_2} {+ \frac{2}{(\rho_1 +\rho_2)(\rho_1^{-1} + \rho_2^{-1} - \Theta(\Psi(\beta_1, \beta_2)))}}. \]}%
Compared with Proposition \ref{prop_dist_transition}, we see that when $\Psi(\beta_1, \beta_2)$ is smaller than $\rho_1^{-1} + \rho_2^{-1}$ (up to a constant multiple), the transfer is positive.
Moreover, the data efficiency ratio quantifies how effective the positive transfer is using MTL.
%In addition to $\rho_1,\rho_2$, the data efficiency ratio also scales with $\Psi(\beta_1, \beta_2)$, the model distance between the tasks.

Next we state Proposition \ref{prop_data_efficiency}, which gives precise upper and lower bounds on the data efficiency ratio for Taskonomy. Its proof can be found in Appendix \ref{app_proof_32}. %In the statement, we shall denote the data efficiency ratio as $\al^\star$.

\begin{proposition}%[Labeled data efficiency]
\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1,\rho_2 \ge 9$ and $\Psi(\beta_1, \beta_2) < (5(\rho_1-1))^{-1} + (5(\rho_2-1))^{-1}$.
	Then the data efficiency ratio $x^\star$ satisfies 
	\be\label{eq_uplowx} x_l  \le x^\star\le \frac{1}{\rho_1 + \rho_2} \bigbrace{  \frac{2}{(\rho_1-1)^{-1} + (\rho_2-1)^{-1} - 5\Psi(\beta_1, \beta_2)}+1}, \ee
	where we denoted
	$$x_l:= \frac1{\rho_1+\rho_2}\left(\frac{2}{(\rho_1-1)^{-1}+(\rho_2 -1)^{-1}}+1\right).$$
\end{proposition}
